Negative and Fractional Types

### 2012-06-28 rational_ext.tex
relations

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! Gx = \Gamma^{\times}
%! G = \Gamma
%! [] = \Box
%! |-->* = \mapsto^{\asterix}
%! |-->> = \mapsto_{\ggg}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! <><> = \approx
%! ==> = \Longrightarrow
%! <== = \Longleftarrow
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! -o+ = \multimap^{+}
%! -o* = \multimap^{\times}
%! -o = \multimap
%! ::= = &::=&
%! /= = \neq
%! @@ = \mu
%! [^ = \lceil
%! ^] = \rceil
%! {| = \{
%! |} = \}
%! dots = \ldots
%! ^^^ = ^{\dagger}
%! elem = \in
%! forall = \forall
%! exists = \exists
%! empty = \epsilon
%! Pi = \Pi
%! Pi0 = \Pi^{o}
%! PiEE* = \Pi^{\eta\epsilon}_{*}
%! PiEE+ = \Pi^{\eta\epsilon}_{+}
%! PiEE = \Pi^{\eta\epsilon}
%! CatSet = \textbf{Set}
%! theseus = Theseus
%! sqrt(x) = \sqrt{#x}
%! surd(p,x) = \sqrt[#p]{#x}
%! inv(x) = \frac{1}{#x}
%! frac(x,y) = \frac{#x}{#y}
%! * = \times

@@

In arithmetic the main identity satisfied by fractional numbers is that for
non-zero {{b}}, we have {{b * 1/b = 1}}. Translating this identity to our
world requires us to find an isomorphism between the types {{1}} and 
{{b * 1/b}} (where {{b}} is not equivalent to {{0}}). Formally, we move 
from the setting of symmetric monoidal categories to that of compact
closed categories.

%%%%%%%%%%%%%%%%%
\subsection{Pointed Sets}

In order to interpret fractional types, we first lift the values and
combinators of {{Pi}} to the world of \emph{pointed sets}. Intuitively:
\begin{itemize}
\item instead of interpreting a type {{b}} as a plain set of cardinality
  $\modulusp{b}$, we now interpret each type {{b}} as a pointed set $P_b$
  with a distinguished \emph{basepoint} $\bot_b$;
\item a combinator {{c : b1 <-> b2}} is interpreted as a function as before
  with the restriction that it must map the basepoint $\bot_{b_1}$ to the
  basepoint $\bot_{b_2}$.
\end{itemize}

Intuitively a pointed set may be seen as a vector space over the field with
one element. This view allows us to add the type {{1/b}} and interpret it as
the \emph{dual} of the vector space denoting {{b}}. Concretely the elements
of the pointed set denoted by {{1/b}} are \emph{linear functionals} $f : P_b
\rightarrow P_1$. Note that $P_1$ is the pointed set containing {{()}} and
$\bot_1$. The isomorphism between the types {{1}} and {{b * 1/b}} is now a
consequence of that fact that the pairing of a vector and a functional in the
dual space produces a scalar.

\newpage

\paragraph*{Data.} The syntax of {{Pi}} is extended with the type {{S b}}
which denotes sets {{ {| v1, v2, dots |} }} of values of type {{b}}:
%subcode{bnf} include main
% value types, b ::= 0 | 1 | b + b | b * b | S b
% values, v ::= () | left v | right v | (v, v) | {| v1, v2, dots |}

\paragraph*{Programs.} All the programs of {{Pi}} are special maps of type
{{v -> v}}. Specifically, they are witnesses to the following type
isomorphisms:
%subcode{bnf} include main
%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
%zeroe :&  0 + b <-> b &: zeroi
%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
%identl* :&  1 * b <-> b &: identr*
%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
%dist0 :& 0 * b <-> 0 &: factor0
%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
Each line of the above table introduces one or two combinators that witness
the isomorphism in the middle. Collectively the isomorphisms state that the
structure {{(b,+,0,*,1)}} is a \emph{commutative semiring}, i.e., that each
of {{(b,+,0)}} and {{(b,*,1)}} is a commutative monoid and that
multiplication distributes over addition. The isomorphisms are extended to
form a congruence relation by adding the following constructors that witness
equivalence and compatible closure:
%subcode{proof} include main
%@  ~
%@@ id : b <-> b 
%
%@ c : b1 <-> b2
%@@ sym c : b2 <-> b1
%
%@ c1 : b1 <-> b2
%@ c2 : b2 <-> b3
%@@ c1(;)c2 : b1 <-> b3
%---
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
%
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4

\begin{definition}{(Syntax of {{Pi}})}
\label{def:Pi}
We collect our types, values, and combinators, to get the full language
definition.
%subcode{bnf} include main
% value types, b ::= 0 | 1 | b+b | b*b 
% values, v ::= () | left v | right v | (v,v) 
%
% comb.~types, t ::= b <-> b
% iso ::= zeroe | zeroi 
%     &|& swap+ | assocl+ | assocr+ 
%     &|& identl* | identr* 
%     &|& swap* | assocl* | assocr* 
%     &|& dist0 | factor0 | dist | factor 
% comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
\end{definition}

%subcode{opsem} include main
% eta* &: 1 <-> (1/b) * b :& eps*

IntuitivelyThe main challenge Adding fractional types to
{{Pi}} takes us from a symmetric monoidal category to a compact closed
category.

>>>>

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! Gx = \Gamma^{\times}
%! G = \Gamma
%! [] = \Box
%! |-->* = \mapsto^{\asterix}
%! |-->> = \mapsto_{\ggg}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! <><> = \approx
%! ==> = \Longrightarrow
%! <== = \Longleftarrow
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! -o+ = \multimap^{+}
%! -o* = \multimap^{\times}
%! -o = \multimap
%! ::= = &::=&
%! /= = \neq
%! @@ = \mu
%! [^ = \lceil
%! ^] = \rceil
%! {| = \{
%! |} = \}
%! bot = \bot
%! bot1 = \bot_1
%! dots = \ldots
%! ^^^ = ^{\dagger}
%! elem = \in
%! forall = \forall
%! exists = \exists
%! empty = \epsilon
%! Pi = \Pi
%! Pi0 = \Pi^{o}
%! PiEE* = \Pi^{\eta\epsilon}_{*}
%! PiEE+ = \Pi^{\eta\epsilon}_{+}
%! PiEE = \Pi^{\eta\epsilon}
%! CatSet = \textbf{Set}
%! theseus = Theseus
%! sqrt(x) = \sqrt{#x}
%! surd(p,x) = \sqrt[#p]{#x}
%! inv(x) = \frac{1}{#x}
%! frac(x,y) = \frac{#x}{#y}
%! * = \times

@@

\begin{verbatim}
revise the paragraph below to 
say something that's consistent with the 
presentation in the previous section; the
previous type system was the natural numbers;
we presented a natural language of terms
for that type system and a semantics in 
the rig category of finite
sets and bijections (two symmetric
monoidal structures and distributivity).
now we extend the type system to correspond
to the non-negative rational numbers;
we want to find the corresponding language
and semantic interpretation
\end{verbatim}

In arithmetic the main identity satisfied by fractional numbers is that for
non-zero {{b}}, we have {{b * 1/b = 1}}. Accommodating this identity in the
setting of {{Pi}} requires us to find an isomorphism between the types {{1}}
and {{b * 1/b}} (where {{b}} is not equivalent to {{0}}). Formally, we move
from the setting of symmetric monoidal categories to that of compact closed
categories.

%%%%%%%%%%%%%%%%%
\subsection{Non-determinism}

In order to interpret fractional types, we first lift the values and
combinators of {{Pi}} to the world of sets and relations. The lifting is
realized using a monad whose functor is powerset. Intuitively:
\begin{itemize}
\item instead of interpreting a type {{b}} as a finite set of values $\{ v_1,
  v_2, \ldots, v_{\modulusp{b}}\}$, we now interpret each type {{b}} as the
  powerset of such values; note that the type {{0}} now denotes
  $\{\emptyset\}$, i.e., it has one element which is the empty set;
  \item a {{Pi}}-value {{v}} of type {{b}} is lifted to a singleton set 
  $\{ v \}$;
\item each combinator {{c : b1 <-> b2}} is lifted to a \emph{symmetric}
  relation $R_c$ defined as follows:
\[\begin{array}{r@{\qquad}c@{\qquad}l}
\{ v_1, v_2, \ldots \} &R_c& \bigcup \{ c(v_i) \} 
\end{array}\]
For example, the combinator {{swap+ : 1+1 <-> 1+1}} is lifted to the relation $R$:
\[\begin{array}{r@{\qquad}c@{\qquad}l}
\emptyset &R& \emptyset \\
\{ \mathit{left}~() \} &R& \{ \mathit{right}~() \} \\
\{ \mathit{right}~() \} &R& \{ \mathit{left}~() \} \\
\{ \mathit{left}~(), \mathit{right}~() \} &R& \{ \mathit{left}~(), \mathit{right}~() \} 
\end{array}\]
\end{itemize}
Because every combinator corresponds to a bijection, the lifted relation is
symmetric.  Formally we have moved from the category of finite sets and
bijections to the category of finite sets and \emph{symmetric relations}.

Now that we have introduced the potential for non-determinism, it is possible
to introduce a type {{1/b}}, values {{1/v}}, and combinators {{eta*}} and
{{eps*}} that witness the following isomorphism:
%subcode{opsem} include main
% eta* &: 1 <-> (1/b) * b :& eps*
where {{b}} is not ``equivalent'' to {{0}} as detailed next.

Intuitively, we view a value {{1/v}} as a \emph{constraint} that is only
satisfied if it is matched with a value {{v'}} that is identical to
{{v}}. The type {{1/b}} is interpreted as the powerset of all the values
{{1/v}} with {{v}} being a {{Pi}}-value of type {{b}}. For example, the type
{{1/(1+1)}} denotes:
\[
\{ 
\emptyset, 
\{ 1/\mathit{left}~() \}, 
\{ 1/\mathit{right}~() \},
\{ 1/\mathit{left}~(), 1/\mathit{right}~() \} 
\}
\]
Of particular interest is that both the type {{0}} and the type {{1/0}}
denote $\{\emptyset\}$.

Given this interpretation of types and values, the realization of 
{{eta* : 1 <-> (1/b) * b}} is the relation $R$: 
\[\begin{array}{r@{\qquad}c@{\qquad}l}
\emptyset &R& \emptyset \\
\{ () \} &R& \{ (1/v,v) ~|~ v \mbox{~is~a~} \Pi\mbox{-value~of~type~} b \}
\end{array}\]
For example, at type $1+1$, the interpretation of {{eta*}} is the relation:
\[\begin{array}{r@{\qquad}c@{\qquad}l}
\emptyset &R& \emptyset \\
\{ () \} &R& \{ (1/\mathit{left}~(),\mathit{left}~()), 
                (1/\mathit{right}~(),\mathit{right}~()) \} 
\end{array}\]
At type $0$, the interpretation of {{eta*}} is:
\[\begin{array}{r@{\qquad}c@{\qquad}l}
\emptyset &R& \emptyset \\
\{ () \} &R& \emptyset 
\end{array}\]
which is not a symmetric relation. Hence the presence of {{eta*}} introduces
some ``non-reversible'' effect and moves us to the category of finite sets 
and (not necessarily symmetric) relations. The interpretation of 
{{eps* : (1/b) * b <-> 1}} is the relation:
\[\begin{array}{r@{\qquad}c@{\qquad}l}
\emptyset &R& \emptyset \\
\{ \{ (1/v,v) ~|~ v \mbox{~is~a~} \Pi\mbox{-value~of~type~} b \} &R& \{ () \} \\
\{ \{ (1/v,v') ~|~ v \mbox{~is~not~identical~to~} v' \} &R& emptyset 
\end{array}\]
For example, at type $0$, we have:
\[\begin{array}{r@{\qquad}c@{\qquad}l}
\emptyset &R& \emptyset 
\end{array}\]
At type $1+1$, we have:
\[\begin{array}{r@{\qquad}c@{\qquad}l}
\emptyset &R& \emptyset \\
\{ \{ (1/\mathit{left}~(),\mathit{left}~()) \} &R& \{ () \} \\
\{ \{ (1/\mathit{left}~(),\mathit{right}~()) \} &R& \emptyset \\
\{ \{ (1/\mathit{right}~(),\mathit{left}~()) \} &R& \emptyset \\
\{ \{ (1/\mathit{right}~(),\mathit{right}~()) \} &R& \{ () \} \\
\{ \{ (1/\mathit{left}~(),\mathit{left}~()), (1/\mathit{left}~(),\mathit{right}~()) \} &R&
  \emptyset \} \\
\{ \{ (1/\mathit{left}~(),\mathit{left}~()), (1/\mathit{right}~(),\mathit{left}~()) \} &R&
  \emptyset \} \\
\{ \{ (1/\mathit{left}~(),\mathit{left}~()), (1/\mathit{right}~(),\mathit{right}~()) \} &R&
  \{ () \} \} \\
\{ \{ (1/\mathit{left}~(),\mathit{right}~()), (1/\mathit{left}~(),\mathit{left}~()) \} &R&
  \emptyset \} \\
\{ \{ (1/\mathit{left}~(),\mathit{right}~()), (1/\mathit{right}~(),\mathit{left}~()) \} &R&
  \emptyset \} \\
\{ \{ (1/\mathit{left}~(),\mathit{right}~()), (1/\mathit{right}~(),\mathit{right}~()) \} &R&
  \emptyset \} \\
\{ \{ (1/\mathit{right}~(),\mathit{left}~()), (1/\mathit{left}~(),\mathit{left}~()) \} &R&
  \emptyset \} \\
\{ \{ (1/\mathit{right}~(),\mathit{left}~()), (1/\mathit{left}~(),\mathit{right}~()) \} &R&
  \emptyset \} \\
\{ \{ (1/\mathit{right}~(),\mathit{left}~()), (1/\mathit{right}~(),\mathit{right}~()) \} &R&
  \emptyset \} \\
\{ \{ (1/\mathit{right}~(),\mathit{right}~()), (1/\mathit{left}~(),\mathit{left}~()) \} &R&
  \{ () \} \} \\
\{ \{ (1/\mathit{right}~(),\mathit{right}~()), (1/\mathit{left}~(),\mathit{right}~()) \} &R&
  \emptyset \} \\
\{ \{ (1/\mathit{right}~(),\mathit{right}~()), (1/\mathit{right}~(),\mathit{left}~()) \} &R&
  \emptyset \} 
\end{array}\]

The category of finite sets and relations is a compact closed category. Each
object has a dual which is itself. The main coherence condition requires us
to check that:
\[\begin{array}{rcl}
b &\rightarrow& b * 1 \\
  &\rightarrow_{\eta^*}& b * (1/b * b) \\
  &\rightarrow& (b * 1/b) * b \\
  &\rightarrow_{\epsilon^*}& 1 * b \\
  &\rightarrow& b
\end{array}\]
For example, let's check one instance of this identity at the type $1+1$ and
the value {{left ()}}:
\[\begin{array}{rcl}
\{ \mathit{left}~() \} &\rightarrow& \{ (\mathit{left}~(), ()) \} \\
  &\rightarrow_{\eta^*}& \{ (\mathit{left}~(),(1/\mathit{left}~(),\mathit{left}~())), 
                            (\mathit{left}~(),(1/\mathit{right}~(),\mathit{right}~())) \} \\
  &\rightarrow& \{ ((\mathit{left}~(),1/\mathit{left}~()),\mathit{left}~())), 
                   ((\mathit{left}~(),1/\mathit{right}~()),\mathit{right}~())) \} \\
  &\rightarrow_{\epsilon^*}& \{ ((),\mathit{left}~()) \} \\
  &\rightarrow& \{ \mathit{left}~() \} 
\end{array}\]
At type $0$, we would get:
\[\begin{array}{rcl}
\emptyset &\rightarrow& \emptyset \\
  &\rightarrow_{\eta^*}& \emptyset \\
  &\rightarrow& \emptyset \\
  &\rightarrow_{\epsilon^*}& \emptyset \\
  &\rightarrow& \emptyset
\end{array}\]

%%%%%%%%%%%%%%%%
\subsection{Syntax}
